---
title: "BCB546 R Assignment Documentation"
author: "Nesreen"
date: "2026-03-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
This document describes the workflow for processing and visualizing maize and teosinte genotype data. We use two input files: `fang_et_al_genotypes.txt` (genotype data) and `snp_position.txt` (SNP position data).

---

## Load Libraries

```{r libraries, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
```

---

## Data Inspection

### Load Data Files

```{r load_data}
fang_et_al_genotypes <- read.table("fang_et_al_genotypes.txt", header = TRUE, sep = "\t")
snp_position <- read.table("snp_position.txt", header = TRUE, sep = "\t")
```

### Inspect Genotype File

```{r inspect_fang}
dim(fang_et_al_genotypes)
head(fang_et_al_genotypes)
str(fang_et_al_genotypes)
summary(fang_et_al_genotypes)
```

### Inspect SNP Position File

```{r inspect_snp}
dim(snp_position)
head(snp_position)
str(snp_position)
summary(snp_position)
```

---

## Data Processing

### Maize Data

```{r maize_processing}
# Filter maize groups
maize <- subset(fang_et_al_genotypes, Group %in% c("ZMMIL", "ZMMLR", "ZMMMR"))

# Transpose genotype data
maize_t <- t(maize[ , 4:ncol(maize)])
maize_t <- as.data.frame(maize_t)

# Fix column and row names
colnames(maize_t) <- maize$Sample_ID
maize_t$SNP_ID <- rownames(maize_t)

# Prepare SNP position file with only needed columns
snp_small <- snp_position[ , c("SNP_ID", "Chromosome", "Position")]

# Merge genotype and position data
merged_maize <- merge(maize_t, snp_small, by = "SNP_ID")

# Fix column order: SNP_ID, Chromosome, Position first
all_cols <- colnames(merged_maize)
sample_cols <- all_cols[!all_cols %in% c("SNP_ID", "Chromosome", "Position")]
merged_maize <- merged_maize[ , c("SNP_ID", "Chromosome", "Position", sample_cols)]

# Save unknown and multiple position files before removing them
unknown_maize <- merged_maize[merged_maize$Chromosome == "unknown", ]
write.table(unknown_maize, file = "output/maize/maize_unknown.txt", sep = "\t", row.names = FALSE)

multiple_maize <- merged_maize[merged_maize$Chromosome == "multiple", ]
write.table(multiple_maize, file = "output/maize/maize_multiple.txt", sep = "\t", row.names = FALSE)

# Remove unknown and multiple chromosomes
merged_maize <- merged_maize[!merged_maize$Chromosome %in% c("unknown", "multiple"), ]

# Convert Position and Chromosome to numeric
merged_maize$Position <- as.numeric(merged_maize$Position)
merged_maize$Chromosome <- as.numeric(merged_maize$Chromosome)

# Remove NAs
merged_maize <- merged_maize[!is.na(merged_maize$Position), ]

# Add Group column for visualization
merged_maize$Group <- "Maize"

# Split by chromosome
chr_list <- split(merged_maize, merged_maize$Chromosome)

# Write increasing files (missing data = ?)
lapply(names(chr_list), function(chr) {
  one_chr <- chr_list[[chr]]
  one_chr <- one_chr[order(one_chr$Position), ]
  write.table(one_chr, 
              file = paste0("output/maize/increasing/maize_chr", chr, "_increasing.txt"),
              sep = "\t", row.names = FALSE)
})

# Write decreasing files (missing data = -)
lapply(names(chr_list), function(chr) {
  one_chr <- chr_list[[chr]]
  one_chr <- one_chr[order(one_chr$Position, decreasing = TRUE), ]
  one_chr[] <- lapply(one_chr, function(x) gsub("\\?", "-", x))
  write.table(one_chr, 
              file = paste0("output/maize/decreasing/maize_chr", chr, "_decreasing.txt"),
              sep = "\t", row.names = FALSE)
})
```

### Teosinte Data

```{r teosinte_processing}
# Filter teosinte groups
teosinte <- subset(fang_et_al_genotypes, Group %in% c("ZMPBA", "ZMPIL", "ZMPJA"))

# Transpose genotype data
teosinte_t <- t(teosinte[ , 4:ncol(teosinte)])
teosinte_t <- as.data.frame(teosinte_t)

# Fix column and row names
colnames(teosinte_t) <- teosinte$Sample_ID
teosinte_t$SNP_ID <- rownames(teosinte_t)

# Merge genotype and position data
merged_teosinte <- merge(teosinte_t, snp_small, by = "SNP_ID")

# Fix column order
all_cols <- colnames(merged_teosinte)
sample_cols <- all_cols[!all_cols %in% c("SNP_ID", "Chromosome", "Position")]
merged_teosinte <- merged_teosinte[ , c("SNP_ID", "Chromosome", "Position", sample_cols)]

# Save unknown and multiple position files
unknown_teosinte <- merged_teosinte[merged_teosinte$Chromosome == "unknown", ]
write.table(unknown_teosinte, file = "output/teosinte/teosinte_unknown.txt", sep = "\t", row.names = FALSE)

multiple_teosinte <- merged_teosinte[merged_teosinte$Chromosome == "multiple", ]
write.table(multiple_teosinte, file = "output/teosinte/teosinte_multiple.txt", sep = "\t", row.names = FALSE)

# Remove unknown and multiple chromosomes
merged_teosinte <- merged_teosinte[!merged_teosinte$Chromosome %in% c("unknown", "multiple"), ]

# Convert Position and Chromosome to numeric
merged_teosinte$Position <- as.numeric(merged_teosinte$Position)
merged_teosinte$Chromosome <- as.numeric(merged_teosinte$Chromosome)

# Remove NAs
merged_teosinte <- merged_teosinte[!is.na(merged_teosinte$Position), ]

# Add Group column for visualization
merged_teosinte$Group <- "Teosinte"

# Split by chromosome
chr_list_teo <- split(merged_teosinte, merged_teosinte$Chromosome)

# Write increasing files (missing data = ?)
lapply(names(chr_list_teo), function(chr) {
  one_chr <- chr_list_teo[[chr]]
  one_chr <- one_chr[order(one_chr$Position), ]
  write.table(one_chr, 
              file = paste0("output/teosinte/increasing/teosinte_chr", chr, "_increasing.txt"),
              sep = "\t", row.names = FALSE)
})

# Write decreasing files (missing data = -)
lapply(names(chr_list_teo), function(chr) {
  one_chr <- chr_list_teo[[chr]]
  one_chr <- one_chr[order(one_chr$Position, decreasing = TRUE), ]
  one_chr[] <- lapply(one_chr, function(x) gsub("\\?", "-", x))
  write.table(one_chr, 
              file = paste0("output/teosinte/decreasing/teosinte_chr", chr, "_decreasing.txt"),
              sep = "\t", row.names = FALSE)
})
```

---

## Data Visualization

### SNPs per Chromosome

```{r snps_per_chromosome}
# Combine maize and teosinte data
combined <- rbind(
  merged_maize[ , c("SNP_ID", "Chromosome", "Position", "Group")],
  merged_teosinte[ , c("SNP_ID", "Chromosome", "Position", "Group")]
)
combined$Chromosome <- factor(combined$Chromosome, levels = c(1,2,3,4,5,6,7,8,9,10))
combined$Position <- as.numeric(combined$Position)

p1 <- ggplot(combined, aes(x = as.factor(Chromosome), fill = Group)) +
  geom_bar(position = "dodge") +
  labs(title = "SNPs per Chromosome", x = "Chromosome", y = "Number of SNPs") +
  theme_bw()
p1
ggsave("figures/SNPs_per_chromosome.png", plot = p1, width = 10, height = 6, dpi = 300)
```

Chromosome 1 has the most SNPs (~155) for both groups while chromosome 10 has the fewest. Maize and teosinte have very similar SNP counts per chromosome.

### Missing Data and Heterozygosity

```{r heterozygosity_maize}
# Make maize data tidy
maize_long <- pivot_longer(merged_maize,
                           cols = -c(SNP_ID, Chromosome, Position, Group),
                           names_to = "Sample",
                           values_to = "Genotype")

# Classify genotypes
maize_long <- maize_long %>%
  mutate(Category = case_when(
    Genotype %in% c("A/A", "C/C", "G/G", "T/T") ~ "Homozygous",
    Genotype == "?/?" ~ "Missing",
    TRUE ~ "Heterozygous"
  ))

p2 <- ggplot(maize_long, aes(x = Sample, fill = Category)) +
  geom_bar(position = "fill") +
  labs(title = "Maize: Proportion of Homozygous, Heterozygous and Missing Data",
       x = "Sample", y = "Proportion") +
  theme_bw() +
  theme(axis.text.x = element_blank())
p2
ggsave("figures/maize_heterozygosity.png", plot = p2, width = 10, height = 6, dpi = 300)
```

```{r heterozygosity_teosinte}
# Make teosinte data tidy
teosinte_long <- pivot_longer(merged_teosinte,
                              cols = -c(SNP_ID, Chromosome, Position, Group),
                              names_to = "Sample",
                              values_to = "Genotype")

# Classify genotypes
teosinte_long <- teosinte_long %>%
  mutate(Category = case_when(
    Genotype %in% c("A/A", "C/C", "G/G", "T/T") ~ "Homozygous",
    Genotype == "?/?" ~ "Missing",
    TRUE ~ "Heterozygous"
  ))

p3 <- ggplot(teosinte_long, aes(x = Sample, fill = Category)) +
  geom_bar(position = "fill") +
  labs(title = "Teosinte: Proportion of Homozygous, Heterozygous and Missing Data",
       x = "Sample", y = "Proportion") +
  theme_bw() +
  theme(axis.text.x = element_blank())
p3
ggsave("figures/teosinte_heterozygosity.png", plot = p3, width = 10, height = 6, dpi = 300)
```

Both maize and teosinte show similar patterns with homozygous sites being dominant (~75-80%).

### Heterozygosity by Group

```{r heterozygosity_by_group}
# Combine maize and teosinte long data
combined_long <- rbind(maize_long, teosinte_long)

p4 <- ggplot(combined_long, aes(x = Group, fill = Category)) +
  geom_bar(position = "fill") +
  labs(title = "Heterozygosity by Group",
       x = "Group", y = "Proportion") +
  theme_bw()
p4
ggsave("figures/heterozygosity_by_group.png", plot = p4, width = 10, height = 6, dpi = 300)
```

### SNP Density Across Chromosomes

```{r snp_density}
p5 <- ggplot(combined, aes(x = Position, fill = Group)) +
  geom_histogram(bins = 50) +
  facet_wrap(~Chromosome, scales = "free_x") +
  labs(title = "SNP Density Across Chromosome Positions",
       x = "Position", y = "Count") +
  theme_bw() +
  theme(axis.text.x = element_blank())
p5
ggsave("figures/SNP_density.png", plot = p5, width = 12, height = 8, dpi = 300)
```

---

## Conclusions

- Chromosome 1 has the highest SNP density for both maize and teosinte
- Maize and teosinte have very similar SNP distributions across chromosomes
- Homozygous sites dominate in all groups (~75-80%)
- ZMMMR has the most missing data while ZMPJA has the least
- ZMMLR and ZMPJA show the highest heterozygosity rates
